---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r}
setwd("~/Coursera/Data Science Specialisation/05 Reproducible Research/Assignment1")
data <- read.csv("activity.csv")
data$date <- as.Date(data$date)
```

## What is mean total number of steps taken per day?

1. Calculate the total number of steps taken per day
```{r}
tapply(data$steps, data$date, sum, na.rm=TRUE)
```

2. Make a histogram of the total number of steps taken each day
```{r}
daily.sum <- tapply(data$steps, data$date, sum, na.rm=TRUE)
barplot(daily.sum, col="skyblue", las=2, main="Total number of steps taken each day")
```

3. Calculate and report the mean and median of the total number of steps taken per day
```{r}
tapply(data$steps, data$date, mean, na.rm=TRUE)
tapply(data$steps, data$date, median, na.rm=TRUE)
```


## What is the average daily activity pattern?

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
data$interval <- as.factor(data$interval)
avg.steps <- tapply(data$steps, data$interval, mean, na.rm=TRUE)
plot(names(avg.steps), avg.steps, type="l", col="red1", xlab="5-minute interval", ylab="Average number of steps taken", main="Number of steps taken per 5-minute interval, averaged across all days")
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
names(which(avg.steps==max(avg.steps)))
```

## Imputing missing values

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r}
sum(is.na(data$steps)) + sum(is.na(data$date)) + sum(is.na(data$interval))
```

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

This strategy uses the mutate function in dplyr to locate missing values and replace them with the mean for that interval. 

```{r}
library(dplyr)
data %>% group_by(interval) %>% mutate(steps = ifelse(is.na(steps), mean(steps, na.rm = T), steps))
```

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
data.imp <- data %>% group_by(interval) %>% mutate(steps = ifelse(is.na(steps), mean(steps, na.rm = T), steps))
```

4. Make a histogram of the total number of steps taken each day

```{r}
barplot(tapply(data.imp$steps, data.imp$date, sum), col="springgreen", las=2, main="Total number of steps taken each day")
```

Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r}
tapply(data.imp$steps, data.imp$date, mean)
tapply(data.imp$steps, data.imp$date, median)
```

These mean and median values are different from the previously calcuated estimates. For days where there was missing data, the imputed values have been substituted which has increased the estimated total daily number of steps for those days. 

## Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r}
data.imp <- mutate(data.imp, day=weekdays(date))
data.imp <- data.imp %>% group_by(day) %>% mutate(weekday = ifelse(day == "Saturday", "weekend", ifelse(day == "Sunday", "weekend", "weekday")))
```

2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

```{r}
weekday.data <- data.imp[(data.imp$weekday=="weekday"), ]
weekend.data <- data.imp[(data.imp$weekday=="weekend"), ]
par(mfrow=c(1,2))
plot(names(tapply(weekday.data$steps, weekday.data$interval, mean)), 
     tapply(weekday.data$steps, weekday.data$interval, mean), 
     type="l", 
     col="darkmagenta",
     main="Average steps per interval on weekdays", 
     xlab="Interval", 
     ylab="Average steps per interval")
plot(names(tapply(weekend.data$steps, weekend.data$interval, mean)), 
     tapply(weekend.data$steps, weekend.data$interval, mean), 
     type="l", 
     col="blue4",
     main="Average steps per interval on weekends", 
     xlab="Interval", 
     ylab="Average steps per interval")
```
